Hydras \& Co.

### 2023-05-12 doc/chapter-fol.tex
work on chapter 12

@@

A \emph{language} is a structure composed of relation and function symbols, each symbol is given an \emph{arity} (number of arguments). As suggested by Russel O'Connor in~\cite{OConnor05}, we consider two arity functions instead of a function defined on the sum type \texttt{Relations + Functions}.

@@

\subsubsection{A toy language}
In order to show a few simple examples of statements and proofs, we define a small language with very few symbols:
two constant symbols: $a$ and $b$, three function symbols $f$, $g$ and $h$  (of respective arity 1,1 and 2), three propositional symbols $A$, $B$ and $C$, a one-place predicate symbol $P$, and a binary relational symbol $R$.

@@

Given a language $L$, we define the set of \emph{terms} and
$n$-\emph{tuples} of terms.

@@

Let us give the \gallina terms of opur toy language, corresponding to the terms $a$,
$f(a)$, and $h(f(a),a)$. The ``term'' $h(a)$ is not well formed, since the arity of $h$ is not respected.

@@

\begin{remark}[Variables]
In O'Connor's formalization of first-order logic, the variables are \gallina terms of the form \texttt{var $L$ $i$}, where $L$ is
 a language and $i$ a natural number. 
Although other choices may be considered : PHOAS, de Bruijn indices, etc,  we   respect this representation, in order not to break long proof scripts which use this representation (please look at \cite{OConnor05} for a related discussion).

\subsubsection{Simplified notations for terms}
\label{sec:fol-term-notations}

 We can use \coq's \texttt{Notation} features to print and parse terms in a more readable form.

\inputsnippets{FolExamples/beforeAbbreviations, FolExamples/termAbbreviations, FolExamples/NowItsBetter}

@@

The full abstract syntax of first-order formulas is completed 
with a few more notations. Note that we are studying \emph{classical} first-order logic, so the existential quantifier is derived from the universal one through negation.

@@

\subsection{Examples}

Let us give a few examples of first-order formulas. on our toy language.

\label{fol:examplesf1f2f3}
\begin{enumerate}
\item $R\;a\;b$
\item $\forall v_0 v_1, (R\;v_0\;v_1 \leftrightarrow R\;v_1\;v_0$
\item $ \forall v_0, v_0=a \vee \exists\;v_1,\, v_0= f(v_1)$
\end{enumerate}

@@

\subsection{Notations for first-order formulas}

In order to make the first order formulas more readable, we use a generic notation scope called \texttt{fol\_scope}, which uses mainly \coq's 
notations for connectives and quantifiers. The \texttt{\%fol} delimiter 
allows the user to distinguisf \texttt{FOL} connectives from their \coq equivalent.

@@

\inputsnippets{fol/folScope}

@@

In addition to \texttt{fol\_scope}, we introduce a few abbreviations specific to our toy language.

@@

The following interactions show how the formulas on $L$ may be 
parsed or printed.

@@

\begin{remark}[bound variables]

@@

\inputsnippets{FolExamples/boundVars}  
\end{remark}

\subsubsection{Languages for Arithmetic}

In ~\href{../theories/html/hydras.Ackermann.Languages.html}{Ackermann.Languages}, we can find two examples of languages built with the usual symbols of arithmetic.

The first language, \texttt{LNT} (\emph{Language of Number Theory}) has just function symbols for $+$, $\times$, $0$ and successor.

The second language, LNN, is the language of NN (\emph{Natural Numbers}) and has
the same function symbols as LNT plus one relation symbol for the strict inequality $<$ : \texttt{LT} (less than).

First, we declare two ranked alphabets.

\begin{todo}
Start with toy Language , then LNT and LNN.
\end{todo}

@@

>>>>

This chapter is devoted to the presentation to  data structures for representing first order formulas over a ranked alphabet, and the basic functions and predicates over these types.
Although all the following constructions come directly from~\cite{Goedel}, we introduced minor changes (mainly syntax) to
take into account recent changes in \coq(new constructions, tactics, notations, etc.).

@@

A \emph{language} is a structure composed of relation and function symbols, each symbol is given an \emph{arity} (number of arguments)\,\footnote{As suggested by Russel O'Connor in~\cite{OConnor05}, we consider two arity functions instead of a function defined on the sum type \texttt{Relations + Functions}}.

@@

\subsubsection{Example: a toy language}
In order to show a few simple examples of statements and proofs, we define a small language with very few symbols:
two constant symbols: $a$ and $b$, three function symbols $f$, $g$ and $h$  (of respective arity 1, 1 and 2), three propositional symbols $A$, $B$ and $C$, a one-place predicate symbol $P$, and a binary relational symbol $R$.

@@

Given a language $L$, we define the type of \emph{terms} and
$n$-\emph{tuples of terms} over $L$.

@@

\begin{remark}[Variables]
In O'Connor's formalization of first-order logic,  variables are \gallina terms of the form (\texttt{var $L$ $i$}), where $L$ is
 a language and $i$ a natural number. 
Although other choices may be considered : PHOAS, de Bruijn indices, etc,  we   respect this representation, in order not to break long proof scripts which use this representation (please look at \cite{OConnor05} for a related discussion).

\begin{remark}
This representation of terms uses mutually inductive data-types instead of list of terms. Please see also Remark~\ref{hydra:mutually-inductive-vs-lists}.
\end{remark}

\subsubsection{Examples}
\label{sect:folTermExamples}

Let us define some \gallina terms over our toy language, respecively corresponding to the terms $a$,
$f(a)$, and $h(f(a),a)$. The fourth ``term'' is not well formed, since the arity of $h$ is not respected\,\footnote{Strictly speaking, it's not a (well typed) term!}.

@@

@@

The  abstract syntax of first-order formulas is completed 
with a few abbreviations\,\footnote{Technically, these abbreviations take the form of \emph{global definitions}}. Note that we are studying \emph{classical} first-order logic, for instance the existential quantifier is derived from the universal one through negation.

@@

\subsection{Examples}
\label{sect:folFormExamples}

Let us give a few examples of first-order formulas over $L$.

\label{fol:examplesf1f2f3}
\begin{enumerate}
\item $R\;a\;b$
\item $\forall v_0\; v_1, R\;v_0\;v_1 \leftrightarrow R\;v_1\;v_0$
\item $ \forall v_0, v_0=a \vee \exists\;v_1,\, v_0= f(v_1)$
\end{enumerate}

@@

\section{Simplified notations for terms and formulas}
\label{sect:fol-notations}

 We use \coq's \texttt{Notation} features to print and parse terms and formulas  in a more readable form.

 \subsection{Parsing and printing terms}
 \label{sec:fol-term-notations}

Let us consider the term $t_2$ of Sect\ref{sect:folTermExamples}~. Its normal form is quite hard to read:

\inputsnippets{FolExamples/beforeAbbreviations}

We may use \coq's abbreviation feature  to print and parse $L$-terms with a more natural concrete syntax.

\inputsnippets{FolExamples/termAbbreviations, FolExamples/NowItsBetter}

\subsection{Parsing and printing first-order formulas}

In order to make the first order formulas more readable, we define  a  notation scope called \texttt{fol\_scope}, which contains 
notations for connectives and quantifiers. The \texttt{\%fol} delimiter 
allows the user to distinguish \texttt{FOL} connectives from their \coq equivalent.

@@

\inputsnippets{fol/folScope1}

The following notations are used (when printing) when  some computation 
    expands a disjunction, conjuction, etc. 
    in terms of implication and negation.

\inputsnippets{fol/folScope2}

@@

In addition to \texttt{fol\_scope}, we can introduce a few abbreviations specific to our toy language.

@@

The following interactions show how the formulas 
of~\ref{sect:folFormExamples} may be 
parsed or printed.

@@

\section{Computing and reasoning on first-order formulas}

\subsection{Bound variables}

@@

\inputsnippets{FolExamples/boundVars}

  \begin{todo}
   Link to the lemmas which attest the equivalence of these formulas (properties of substitution, logical equivalence).
  \end{todo}

@@

\section{Languages for Arithmetic}

Two languages built with the usual symbols of arithmetic are 
defined in ~\href{../theories/html/hydras.Ackermann.Languages.html}{Ackermann.Languages}.

\begin{itemize}
\item The first language: \texttt{LNT} (\emph{Language of Number Theory}) has just function symbols for $+$, $\times$, $0$ and successor.
\item The second language: LNN (\emph{Language of Natural Numbers})  has
the same function symbols as LNT plus one relation symbol for the strict inequality $<$ : \texttt{LT} (less than).
\end{itemize}

\subsubsection{Language of Number Theory (LNT)}

First, we declare two alphabets.

\subsubsection{Language of Natural Numbers (LNN)}

\texttt{LNN} is an extension of \texttt{LNT}, by the addition 
of the $<$ relation symbol.

\inputsnippets{Languages/LNNDef}
